```{r data_import}
// ... existing data import code ...
```

## Study Population and Sample Selection

```{r flow_diagram, fig.width=10, fig.height=12}
# Load required packages
if(!require(DiagrammeR)) {
  install.packages("DiagrammeR")
  library(DiagrammeR)
}

# Calculate counts for flow diagram
initial_counts <- anonymized_data %>%
  summarise(
    total_samples = n(),
    total_patients = n_distinct(anonymous_id)
  )

genotype_counts <- anonymized_data %>%
  group_by(genotype) %>%
  summarise(
    samples = n(),
    patients = n_distinct(anonymous_id)
  )

# Calculate assay-specific counts
assay_counts <- anonymized_data %>%
  summarise(
    advia_samples = sum(!is.na(hb_advia)),
    lorrca_samples = sum(!is.na(ei_min_lorrca)),
    fcell_samples = sum(!is.na(percent_f_cells)),
    dense_rbc_samples = sum(!is.na(dense_rbcs)),
    viscosity_samples = sum(!is.na(visc_45))
  )

# Calculate age group counts
age_counts <- anonymized_data %>%
  group_by(age_group) %>%
  summarise(
    samples = n(),
    patients = n_distinct(anonymous_id)
  )

# Create the flow diagram
flow_diagram <- grViz("
  digraph flow {
    # Graph settings
    graph [rankdir = TB, splines = ortho]
    node [shape = rectangle, style = filled, fillcolor = white, fontname = Helvetica, fontsize = 12]
    
    # Nodes
    initial [label = 'Initial SCD Database\n(n = samples)', width = 3]
    
    genotype [label = 'Genotype Filter\nHbSS and HbSC only', width = 3]
    
    ss [label = 'HbSS\nSamples: XX\nPatients: YY', width = 2]
    sc [label = 'HbSC\nSamples: XX\nPatients: YY', width = 2]
    
    assays [label = 'Available Assay Data\n(samples with results)', width = 4]
    
    advia [label = 'Complete Blood Count\nSamples: XX', width = 2]
    lorrca [label = 'LORRCA\nSamples: XX', width = 2]
    fcell [label = 'F-Cells\nSamples: XX', width = 2]
    dense [label = 'Dense RBCs\nSamples: XX', width = 2]
    visc [label = 'Viscosity\nSamples: XX', width = 2]
    
    age [label = 'Age Group Distribution', width = 3]
    
    young [label = '18-45 years\nSamples: XX\nPatients: YY', width = 2]
    middle [label = '46-59 years\nSamples: XX\nPatients: YY', width = 2]
    elderly [label = '60+ years\nSamples: XX\nPatients: YY', width = 2]
    
    # Edges
    initial -> genotype
    genotype -> {ss sc}
    {ss sc} -> assays
    assays -> {advia lorrca fcell dense visc}
    {advia lorrca fcell dense visc} -> age
    age -> {young middle elderly}
    
    # Edge settings
    edge [color = black, penwidth = 1]
  }
  [Initial Sample Count: total_samples = ', initial_counts$total_samples, 
   '\nInitial Patient Count: total_patients = ', initial_counts$total_patients, ']
  [SS Samples: ', genotype_counts$samples[genotype_counts$genotype == 'SS'],
   '\nSS Patients: ', genotype_counts$patients[genotype_counts$genotype == 'SS'], ']
  [SC Samples: ', genotype_counts$samples[genotype_counts$genotype == 'SC'],
   '\nSC Patients: ', genotype_counts$patients[genotype_counts$genotype == 'SC'], ']
  [Advia: ', assay_counts$advia_samples, ']
  [LORRCA: ', assay_counts$lorrca_samples, ']
  [F-Cells: ', assay_counts$fcell_samples, ']
  [Dense RBCs: ', assay_counts$dense_rbc_samples, ']
  [Viscosity: ', assay_counts$viscosity_samples, ']
  [18-45 Samples: ', age_counts$samples[age_counts$age_group == '18-45'],
   '\n18-45 Patients: ', age_counts$patients[age_counts$age_group == '18-45'], ']
  [46-59 Samples: ', age_counts$samples[age_counts$age_group == '46-59'],
   '\n46-59 Patients: ', age_counts$patients[age_counts$age_group == '46-59'], ']
  [60+ Samples: ', age_counts$samples[age_counts$age_group == '60+'],
   '\n60+ Patients: ', age_counts$patients[age_counts$age_group == '60+'], ']
")

# Display the flow diagram
flow_diagram

# Create a summary table of sample counts
sample_summary <- bind_rows(
  # Initial counts
  initial_counts %>% 
    mutate(Category = "Initial Database", .before = everything()) %>%
    rename(Samples = total_samples, Patients = total_patients),
  
  # Genotype counts
  genotype_counts %>%
    mutate(Category = paste("Genotype:", genotype)) %>%
    select(-genotype),
  
  # Age group counts
  age_counts %>%
    mutate(Category = paste("Age Group:", age_group)) %>%
    select(-age_group)
)

# Display the summary table
kable(sample_summary,
      caption = "Sample and Patient Counts by Category",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

# Create a summary table of assay counts
assay_summary <- pivot_longer(assay_counts, 
                            everything(),
                            names_to = "Assay",
                            values_to = "Samples") %>%
  mutate(
    Assay = case_when(
      Assay == "advia_samples" ~ "Complete Blood Count",
      Assay == "lorrca_samples" ~ "LORRCA",
      Assay == "fcell_samples" ~ "F-Cells",
      Assay == "dense_rbc_samples" ~ "Dense RBCs",
      Assay == "viscosity_samples" ~ "Viscosity",
      TRUE ~ Assay
    )
  )

# Display the assay summary table
kable(assay_summary,
      caption = "Sample Counts by Assay Type",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

The flow diagram above illustrates our study population selection process and available data for each assay type. Starting from our initial SCD database, we applied the following criteria:

1. **Genotype Selection**: 
   - Included only HbSS and HbSC patients
   - Excluded other genotypes (HbSβ⁰, HbSβ⁺, etc.)

2. **Assay Availability**:
   - Complete Blood Count (CBC) from Advia analyzer
   - LORRCA rheology measurements
   - F-Cell percentages
   - Dense RBC percentages
   - Blood viscosity measurements

3. **Age Group Stratification**:
   - Young adults (18-45 years)
   - Middle-aged adults (46-59 years)
   - Elderly adults (60+ years)

The tables above provide detailed counts of samples and unique patients for each category and assay type. These numbers reflect our final analytic dataset used for all subsequent analyses. 