---
title: "EPIC Dataset Overview"
author: "Jonathan Wade"
format:
  html:
    code-fold: true
  pdf:
    documentclass: article
    papersize: letter
    toc: true
    number-sections: true
    colorlinks: true
    fig-width: 7
    fig-height: 4
    fig-pos: 'H'
    geometry:
      - margin=1in
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(lubridate)
library(janitor)
library(knitr)
library(cowplot)
library(kableExtra)

theme_set(theme_cowplot())
```

## 1. Dataset Overview

```{r load-data}
data_path <- Sys.getenv("DATA_DIR", unset = "./data/epic/")

# Read all files
file_list <- list.files(data_path, pattern = "\\.txt$", full.names = TRUE)
data_list <- map(file_list, ~ read_delim(.x, delim = "|"))
names(data_list) <- basename(file_list)

# Clean all dataframes immediately after loading
clean_data_list <- map(data_list, clean_names)

# Create summary table of records and patients per file using cleaned data
file_summary <- map_dfr(names(clean_data_list), ~{
  df <- clean_data_list[[.x]]
  tibble(
    File_Name = .x,
    Total_Records = nrow(df),
    Unique_Patients = n_distinct(df$patient_mrn),
    Number_of_Variables = ncol(df)
  )
})

kable(file_summary, caption = "Overview of Dataset Files")
```


## 2. Variable Inventory

```{r variable-inventory}
# Function to get variable information
get_var_info <- function(df) {
  map_dfr(names(df), ~{
    example_vals <- if(is.numeric(df[[.x]])) {
      paste(range(df[[.x]], na.rm = TRUE), collapse = " to ")
    } else {
      # Truncate string values to first 50 characters
      vals <- head(unique(df[[.x]]), 3)
      vals <- substr(vals, 1, 50)
      paste(vals, collapse = ", ")
    }
    
    tibble(
      Variable_Name = .x,
      Data_Type = class(df[[.x]])[1],
      N_Missing = sum(is.na(df[[.x]])),
      N_Unique = n_distinct(df[[.x]], na.rm = TRUE),
      Example_Values = example_vals
    )
  })
}

# Create variable inventory for each file
var_inventory <- map(clean_data_list, get_var_info)

# Create individual tables for each file
bone_marrow_vars <- kable(var_inventory[["IN495734_SCD_Bone_Marrow_Baseline.txt"]], 
                         caption = "Variables in Bone Marrow Dataset",
                         format = "html",
                         table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

demographics_vars <- kable(var_inventory[["IN495734_SCD_Demographics_Social_Hx_Baseline.txt"]], 
                         caption = "Variables in Demographics Dataset",
                         format = "html",
                         table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

ip_admissions_vars <- kable(var_inventory[["IN495734_SCD_IP_Admissions_Baseline.txt"]], 
                          caption = "Variables in Inpatient Admissions Dataset",
                          format = "html",
                          table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

ip_meds_vars <- kable(var_inventory[["IN495734_SCD_IP_Meds_Baseline.txt"]], 
                     caption = "Variables in Inpatient Medications Dataset",
                     format = "html",
                     table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

labs_vars <- kable(var_inventory[["IN495734_SCD_Labs_Baseline.txt"]], 
                  caption = "Variables in Laboratory Dataset",
                  format = "html",
                  table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

op_avs_meds_vars <- kable(var_inventory[["IN495734_SCD_OP_AVS_Meds_Baseline.txt"]], 
                         caption = "Variables in Outpatient AVS Medications Dataset",
                         format = "html",
                         table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

op_visits_vars <- kable(var_inventory[["IN495734_SCD_OP_Visits_Baseline.txt"]], 
                       caption = "Variables in Outpatient Visits Dataset",
                       format = "html",
                       table.attr = "style='width:100%;'") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Display all tables
bone_marrow_vars
demographics_vars
ip_admissions_vars
ip_meds_vars
labs_vars
op_avs_meds_vars
op_visits_vars
```

## 3. Demographic Distributions

```{r demographics}
# Clean demographics data
demographics <- clean_data_list$"IN495734_SCD_Demographics_Social_Hx_Baseline.txt" %>%
  mutate(
    birth_date = as_datetime(birth_date),
    age = as.numeric(age)
  )

# Age Distribution
p1 <- ggplot(demographics, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue") +
  labs(title = "Age Distribution", x = "Age", y = "Count")

# Gender Distribution
p2 <- ggplot(demographics, aes(x = gender)) +
  geom_bar(fill = "darkgreen") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count")

# Race Distribution
p3 <- ggplot(demographics, aes(x = race)) +
  geom_bar(fill = "darkred") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Race Distribution", x = "Race", y = "Count")

# Print plots
print(p1)
print(p2)
print(p3)
```

## 4. VOE and ACS Patient Counts

```{r voe-acs}
# Clean admissions data and count VOE/ACS
admissions <- clean_data_list$"IN495734_SCD_IP_Admissions_Baseline.txt" %>%
  mutate(
    has_voe = if_any(
      starts_with("admit_dx_cd_") | starts_with("final_dx_cd_"),
      ~ grepl("^D57\\.0[0-9]", .)
    ),
    has_acs = if_any(
      starts_with("admit_dx_cd_") | starts_with("final_dx_cd_"),
      ~ grepl("^D57\\.01", .)
    )
  )

# Summary of unique patients
patient_summary <- admissions %>%
  group_by(patient_mrn) %>%
  summarise(
    has_any_voe = any(has_voe & !has_acs, na.rm = TRUE),
    has_any_acs = any(has_acs, na.rm = TRUE),
    has_other = any(!has_voe & !has_acs, na.rm = TRUE)
  ) %>%
  summarise(
    Unique_VOE_Patients = sum(has_any_voe),
    Unique_ACS_Patients = sum(has_any_acs),
    Unique_Other_Patients = sum(has_other),
    Total_Unique_Patients = n()
  )

# Summary of total admissions
admission_summary <- admissions %>%
  summarise(
    VOE_Admissions = sum(has_voe & !has_acs, na.rm = TRUE),
    ACS_Admissions = sum(has_acs, na.rm = TRUE),
    Other_Admissions = sum(!has_voe & !has_acs, na.rm = TRUE),
    Total_Admissions = n()
  )

# Combine summaries for display
combined_summary <- bind_rows(
  patient_summary %>% 
    pivot_longer(everything(), names_to = "Metric", values_to = "Count") %>%
    mutate(Type = "Unique Patients"),
  admission_summary %>%
    pivot_longer(everything(), names_to = "Metric", values_to = "Count") %>%
    mutate(Type = "Total Admissions")
)

kable(combined_summary, caption = "VOE and ACS Summary - Patients and Admissions")

# Visualize unique patients
patient_types <- patient_summary %>%
  select(-Total_Unique_Patients) %>%
  pivot_longer(everything(), 
              names_to = "Type", 
              values_to = "Count",
              names_prefix = "Unique_") %>%
  mutate(Type = str_remove(Type, "_Patients"))

# Plot 1: Unique Patients
p1 <- ggplot(patient_types, aes(x = reorder(Type, Count), y = Count)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(
    title = "Distribution of Unique Patients by Crisis Type",
    x = "Crisis Type",
    y = "Number of Unique Patients"
  ) +
  theme_cowplot()

# Plot 2: Total Admissions
admission_types <- admission_summary %>%
  select(-Total_Admissions) %>%
  pivot_longer(everything(), 
              names_to = "Type", 
              values_to = "Count") %>%
  mutate(Type = str_remove(Type, "_Admissions"))

p2 <- ggplot(admission_types, aes(x = reorder(Type, Count), y = Count)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() +
  labs(
    title = "Distribution of Total Admissions by Crisis Type",
    x = "Crisis Type",
    y = "Number of Admissions"
  ) +
  theme_cowplot()

# Print both plots
print(p1)
print(p2)

# Calculate and display average admissions per patient
avg_admissions <- admissions %>%
  group_by(patient_mrn) %>%
  summarise(
    VOE_count = sum(has_voe & !has_acs, na.rm = TRUE),
    ACS_count = sum(has_acs, na.rm = TRUE),
    Other_count = sum(!has_voe & !has_acs, na.rm = TRUE)
  ) %>%
  summarise(
    Avg_VOE_per_patient = mean(VOE_count),
    Avg_ACS_per_patient = mean(ACS_count),
    Avg_Other_per_patient = mean(Other_count)
  )

kable(avg_admissions, caption = "Average Number of Admissions per Patient by Crisis Type")
```

> Note: The first plot shows the number of unique patients who experienced each type of crisis, while the second plot shows the total number of admissions for each crisis type. A patient may be counted in multiple categories if they experienced different types of crises across different admissions.

## 5. Temporal Distribution of Records

```{r temporal-distribution}
# Function to extract dates from different files
get_dates <- function(df, date_col) {
  df %>%
    pull(!!sym(date_col)) %>%
    as_datetime() %>%
    as.Date()
}

# Get dates from different files
admission_dates <- get_dates(admissions, "adm_date_time")
lab_dates <- get_dates(clean_data_list$"IN495734_SCD_Labs_Baseline.txt", "result_time")
op_visit_dates <- get_dates(clean_data_list$"IN495734_SCD_OP_Visits_Baseline.txt", "visit_date")

# Create timeline data
timeline_data <- bind_rows(
  tibble(date = admission_dates, type = "Inpatient Admissions"),
  tibble(date = lab_dates, type = "Lab Results"),
  tibble(date = op_visit_dates, type = "Outpatient Visits")
) %>%
  mutate(type = factor(type, levels = c("Lab Results", "Outpatient Visits", "Inpatient Admissions")))

# Create faceted timeline plot
ggplot(timeline_data, aes(x = date)) +
  geom_histogram(bins = 50, fill = "steelblue") +
  facet_wrap(~ type, ncol = 1, scales = "free_y") +
  labs(
    title = "Timeline of Records by Type",
    subtitle = "Note: Y-axes are scaled independently for each record type",
    x = "Date",
    y = "Count"
  ) +
  theme_cowplot() +
  theme(
    strip.background = element_rect(fill = "lightgray"),
    strip.text = element_text(face = "bold")
  )

# Print summary statistics
record_counts <- timeline_data %>%
  group_by(type) %>%
  summarise(
    Total_Records = n(),
    Earliest_Date = min(date),
    Latest_Date = max(date),
    Avg_Records_per_Month = n() / (as.numeric(diff(range(date)))/30)
  )

kable(record_counts, 
      caption = "Summary of Record Counts and Frequencies",
      digits = 1)
```

> Note: The plots are shown with different y-axis scales to better visualize the patterns within each record type. Lab results typically occur much more frequently than visits or admissions.

## 6. Lab Test Frequencies

```{r lab-frequencies}
# Analyze lab test frequencies
lab_frequencies <- clean_data_list$"IN495734_SCD_Labs_Baseline.txt" %>%
  count(lab_component_description, sort = TRUE) %>%
  head(30)

ggplot(lab_frequencies, aes(x = reorder(lab_component_description, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  coord_flip() +
  labs(title = "Top 30 Most Frequent Lab Tests", x = "Lab Test", y = "Count") +
  theme(axis.text.y = element_text(size = 8))
``` 

## 7. Medication Overview

```{r medication-overview}
# Analyze IP medications
ip_med_summary <- clean_data_list$"IN495734_SCD_IP_Meds_Baseline.txt" %>%
  group_by(rx_class_name) %>%
  summarise(
    n_patients = n_distinct(patient_mrn),
    n_administrations = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(n_patients)) %>%
  head(20)

# Analyze OP medications
op_med_summary <- clean_data_list$"IN495734_SCD_OP_AVS_Meds_Baseline.txt" %>%
  group_by(generic_description) %>%
  summarise(
    n_patients = n_distinct(patient_mrn),
    n_prescriptions = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(n_patients)) %>%
  head(20)

# Visualize IP medications
p1 <- ggplot(ip_med_summary, aes(x = reorder(rx_class_name, n_patients), y = n_patients)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Top 20 Inpatient Medication Classes",
    x = "Medication Class",
    y = "Number of Patients"
  ) +
  theme(axis.text.y = element_text(size = 8))

# Visualize OP medications
p2 <- ggplot(op_med_summary, aes(x = reorder(generic_description, n_patients), y = n_patients)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  coord_flip() +
  labs(
    title = "Top 20 Outpatient Medications",
    x = "Medication",
    y = "Number of Patients"
  ) +
  theme(axis.text.y = element_text(size = 8))

print(p1)
print(p2)
```

## 8. Patient Visit Patterns

```{r visit-patterns}
# Analyze visit frequencies per patient
visit_patterns <- admissions %>%
  group_by(patient_mrn) %>%
  summarise(
    n_admissions = n(),
    avg_days_between = mean(diff(adm_date_time), na.rm = TRUE),
    first_admission = min(adm_date_time),
    last_admission = max(adm_date_time),
    total_days_observed = as.numeric(difftime(max(adm_date_time), 
                                            min(adm_date_time), 
                                            units = "days"))
  ) %>%
  mutate(
    admissions_per_year = (n_admissions / total_days_observed) * 365.25
  )

# Visualize distribution of admissions per patient
p1 <- ggplot(visit_patterns, aes(x = n_admissions)) +
  geom_histogram(binwidth = 1, fill = "steelblue") +
  labs(
    title = "Distribution of Number of Admissions per Patient",
    x = "Number of Admissions",
    y = "Count of Patients"
  )

# Visualize admissions per year
p2 <- ggplot(visit_patterns, aes(x = admissions_per_year)) +
  geom_histogram(binwidth = 0.5, fill = "darkred") +
  labs(
    title = "Distribution of Admissions per Year",
    x = "Admissions per Year",
    y = "Count of Patients"
  )

print(p1)
print(p2)

# Summary statistics
visit_summary <- visit_patterns %>%
  summarise(
    Median_Admissions = median(n_admissions),
    Mean_Admissions = mean(n_admissions),
    Median_Days_Between = median(avg_days_between, na.rm = TRUE),
    Median_Admissions_per_Year = median(admissions_per_year),
    Patients_with_Multiple_Admissions = sum(n_admissions > 1),
    Pct_with_Multiple_Admissions = mean(n_admissions > 1) * 100
  )

kable(visit_summary, caption = "Summary of Patient Visit Patterns", digits = 1)
```

## 9. Data Completeness Analysis

```{r data-completeness}
# Function to analyze missing data
analyze_missing <- function(df) {
  map_dfr(names(df), ~{
    tibble(
      variable = .x,
      n_missing = sum(is.na(df[[.x]])),
      pct_missing = (n_missing / nrow(df)) * 100
    )
  }) %>%
    arrange(desc(pct_missing))
}

# Analyze missing data for each file
missing_data <- map(clean_data_list, analyze_missing)

# Create visualizations for files with missing data
for(file_name in names(missing_data)) {
  missing_df <- missing_data[[file_name]] %>%
    filter(pct_missing > 0)  # Only show variables with missing data
  
  if(nrow(missing_df) > 0) {
    p <- ggplot(missing_df, aes(x = reorder(variable, pct_missing), y = pct_missing)) +
      geom_bar(stat = "identity", fill = "orange") +
      coord_flip() +
      labs(
        title = paste("Missing Data in", file_name),
        x = "Variable",
        y = "Percent Missing (%)"
      )
    print(p)
  }
}
```

## 10. Patient Care Patterns

```{r care-patterns}
# Get unique patients from each type of visit
ip_patients <- clean_data_list$"IN495734_SCD_IP_Admissions_Baseline.txt" %>%
  distinct(patient_mrn)
op_patients <- clean_data_list$"IN495734_SCD_OP_Visits_Baseline.txt" %>%
  distinct(patient_mrn)
lab_patients <- clean_data_list$"IN495734_SCD_Labs_Baseline.txt" %>%
  distinct(patient_mrn)

# Create sets for visualization
library(ggvenn)  # You might need to install this package

patient_sets <- list(
  "Inpatient" = ip_patients$patient_mrn,
  "Outpatient" = op_patients$patient_mrn,
  "Lab Tests" = lab_patients$patient_mrn
)

# Plot Venn diagram
ggvenn(patient_sets, 
       fill_color = c("red", "blue", "green"),
       stroke_size = 0.5,
       set_name_size = 4)
```
